---
title: "pathsimulation_1e5"
author: "Katie Florko"
date: "13/07/2023"
output: html_document
---

We will first estimate the SSUDs from the simple SSF without movement-related covariates. 

```{r}
# generate availability sample
set.seed(2023)
data_ssf <- seal %>%
  mutate(date = as.POSIXct(date)) %>%
  make_track(lon, lat, date) %>%
  steps() %>%
  random_steps() %>% 
  arrange(case_) %>%
  amt::extract_covariates(fish_raster, where = "both") 

# fit SSF1 model
m1 <- data_ssf |> 
  fit_clogit(case_ ~ preydiv_end +  cos(ta_) + log(sl_) + 
               strata(step_id_))
```

We will now simulate a track and visually observe it. Our simulation in this tutorial has 10,000 locations and takes about five minutes to run, however, more steps are recommended (Signer et al. 2017). We kept this tutorial with 10,000 locations due to computation demands. The SSUDs in our paper were generated using include 100,000 locations.

Note that with out track, since we kept our simualation within the study area, we have a slight boundary effect, where more points are present along the boundary of the study area because the simulation bounces back from the edges. 
```{r}
# set starting position for the simulation
start <- make_start((seal %>%
                       mutate(date = as.POSIXct(date)) %>%
                       make_track(lon, lat, date))[1,]) 

# set constants
n_steps = 1e5 # number of steps
n_steps1 = n_steps + 1 # number of steps + 1
burnin <- n_steps/50 # number of steps to remove for the burn-in

# generate redistribution kernel
k1 <- redistribution_kernel(m1, map = fish_raster, start = start,
                            stochastic = TRUE,
                            tolerance.outside = 1,
                            n.control = n_steps)

# simulate path
set.seed(2023)
p1 <- amt::simulate_path(x = k1, n = n_steps, start = start, verbose = TRUE) # this takes about five minutes. We have saved the simulated path to the data folder in the repo; feel free to load it here :)
#p1 <- read.csv(here("data/p1.csv")) %>% as_tibble()


# burn-in
p1_burnt <- p1 %>% dplyr::slice(-c(1:burnin))

# plot simulated track
ssf_track_1 <- fishmap +
  geom_polygon(data = nat_trans, aes(x=long,y=lat,group=group), fill = "grey80", color = "white") +
  geom_point(data = p1_burnt, aes(x = x_,y = y_), alpha = 0.61) +
  geom_path(data = p1_burnt, aes(x = x_,y = y_)) +
  theme_void()
ssf_track_1
```

We can see that it mostly stays within the study area. We will use this track to estimate the SSUD, and visualize the results.
```{r}
# estimate SSUD
uds_ssf1 <- tibble(rep = 1:(n_steps1-burnin), 
    x_ = p1_burnt$x_, y_ = p1_burnt$y_,
    t_ = p1_burnt$t_, dt = p1_burnt$dt) |> 
  filter(!is.na(x_)) |> 
  make_track(x_, y_) |> 
  hr_kde(trast = fish_raster, which_min = "global") %>%
  hr_ud() %>% 
  terra::as.data.frame(xy = TRUE)

# plot SSUD
map_ssf1 <- ggplot() + 
  geom_tile(data = uds_ssf1, aes(x = x,y = y, fill = preydiv)) +
  scale_fill_viridis(option = "mako", name = "SSF1 prediction") +
  geom_polygon(data = nat_trans, aes(x=long,y=lat,group=group), fill = "grey80", color = "white") +
  coord_cartesian(xlim = c(150000,550000), ylim = c(-500000,-50000), expand = F) +
  theme_void()
map_ssf1
```


We will follow the same steps to generate a SSUD for the SSF with movement-related covariates.
```{r}
# fit SSF2 model
m2 <- data_ssf |> 
  fit_clogit(case_ ~ preydiv_end*log(sl_) +  cos(ta_) + strata(step_id_))


# set starting position for the simulation
set.seed(2023)
start <- make_start((seal %>%
                       mutate(date = as.POSIXct(date)) %>%
                       make_track(lon, lat, date))[1,])

# generate redistribution kernel
k2 <- redistribution_kernel(m2, map = fish_raster, start = start,
                            stochastic = TRUE, 
                            tolerance.outside = 1,
                            n.control = n_steps)

# simulate path
set.seed(2023)
p2 <- amt::simulate_path(x = k2, n = n_steps, start = start, verbose = TRUE) # this takes about five minutes. We have saved the simulated path to the data folder in the repo; feel free to load it here :)
#p2 <- read.csv(here("data/p2.csv")) %>% as_tibble()


# burn-in
p2_burnt <- p2 %>% dplyr::slice(-c(1:burnin))

# plot simulated track
ssf_track_2 <- fishmap +
  geom_polygon(data = nat_trans, aes(x=long,y=lat,group=group), fill = "grey80", color = "white") +
  geom_point(data = p2_burnt, aes(x = x_,y = y_), alpha = 0.61) +
  geom_path(data = p2_burnt, aes(x = x_,y = y_)) +
  theme_void()
ssf_track_2


# estimate SSUD
uds_ssf2 <- tibble(rep = 1:(n_steps1-burnin), 
    x_ = p2_burnt$x_, y_ = p2_burnt$y_,
    t_ = p2_burnt$t_, dt = p2_burnt$dt) |> 
  filter(!is.na(x_)) |> 
  make_track(x_, y_) |> 
  hr_kde(trast = fish_raster, which_min = "local") %>%
  hr_ud() %>% 
  terra::as.data.frame(xy = TRUE)

# plot SSUD
map_ss2 <- ggplot() + 
  geom_tile(data = uds_ssf2, aes(x = x,y = y, fill = preydiv)) +
  scale_fill_viridis(option = "mako", name = "SSF2 prediction") +
  geom_polygon(data = nat_trans, aes(x=long,y=lat,group=group), fill = "grey80", color = "white") +
  coord_cartesian(xlim = c(150000,550000), ylim = c(-500000,-50000), expand = F) +
  theme_void()
map_ss2

```
